---
title: "Heatmaps en R"
author: "Midhely García"
date: "3/9/2022"
output: html_document
---
Laboratorio 36 - MAPA DE CALOR -TÉRMICO- with pheatmap
DATOS GENETICOS TOMADOS DE Sahir Bhatnagar.
PRÁCTICA DE CODERS

Objetivo: Realizar un heatmap con datos genéticos 
----------------------------------------------------------------------------------------------------
En este ejercicio vamos a:
1. Cargar nuestra matriz hipotética de datos y dataframes adicionales
2. Realizar varios heatmaps

Un mapa de calor es una representación gráfica de datos que utiliza un sistema de codificación de colores para representar diferentes valores


Heatmaps with pheatmap 
Simulated data created by Sahir Bhatnagar.

possible data pre-processing - normalization - quantile, median, etc., log transform
not necessary here - we have log fold change data that has already been normalized

Calculating your distance matrix (see dist objects):
compute how similar or different you values are
parametric - distance measures based on Pearson correlation 
non parametric - spearman rank - replace by ranks and calculate correlation, Kendall's - relative ordering euclidean - shortest distance between values (has to be normalized), takes magnitude into account city block/Manhattan - sum of distances along each dimension distance 1-correlation - of all pairs of items to be clustered

Cluster your samples (see hclust objects):
hierarchical, organizes into a tree structure based on similarity - short branches if similar and longer branches as similarity decreases repeated cycles where the 2 closest remaining items (smallest distance) get joined by a branch with the length of the branch reflecting the distance between them, the distance between this item and all other remaining items are computed until only one object remains single linkage clustering - distance between 2 items is the minimum of all pairwise distances between items contained in x and y - fast b/c no other calculations need to be performed once you have your distance matrix complete linkage is the maximum of all paiwise distances between x and y average linkage - mean of all pairwise distances between items contained in x and y k-means organize into clusters (self-chosen number) - items are randomly assigned to a cluster - the mean vector fo rall items in each hcluster is computed, items are reassigned to the cluster whose center is closest to them - random starting points so will not always get the same answer, number of trial done to deal with the randomness self organizing maps


Instalar paquetería y llamara a librería
install.packages("pheatmap")
library(pheatmap)

Importar datos utilizando el comando file.choose() para saber su ubicación
```{r}
genes <- as.matrix(
  read.csv("/Users/midhely/Desktop/BMGO_LAB36/INPUT/heatmap_data.csv",
           sep = "," ,
           header = T,
           row.names = 1))

annotation_col <- read.csv("/Users/midhely/Desktop/BMGO_LAB36/INPUT/annotation_col.csv", header = T,
           row.names = 1)

annotation_row <- read.csv("/Users/midhely/Desktop/BMGO_LAB36/INPUT/annotation_row.csv", header = T,
                           row.names = 1)
```

Plotting with pheatmap
```{r}
library(pheatmap)
pheatmap(genes)

```

change font
```{r}
pheatmap(genes, frontsize = 6)
```
default is clustering rows and columns



cluster by gene - groups of similar genes----LOS GENES ESTAN EN LOS RENGLONES
POR DEFAULT CLUSTEA LOS RENGLONES
```{r}
pheatmap(genes, frontsize = 6, cluster_rows = T, cluster_cols = F)
```

cluster by patient - groups of similar patients 
DEBES HACER QUE LAS COLUMNAS SE TRANFOMEN A RENGLONES
```{r}
pheatmap(genes, frontsize = 6, cluster_rows = F, cluster_cols = T)
```

usually order by both
```{r}
pheatmap(genes, frontsize = 6, cluster_rows = T, cluster_cols = T)
```

seeing some patterns emerge - but what do they mean? Great time to add annotation to our plot
```{r}
pheatmap(genes, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row)
```

add to row first, see that genes are clustering according to the pathways they belong to
```{r}
pheatmap(genes, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col)
```
now have information about the drug and condition 

GRAFICO COMPLETO G1
```{r}
pheatmap(genes, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica")
```

GRAFICO QUITANDO CLUSTERS (ARBOLES DE AGRUPACIÓN-DENDOGRAMAS)


take a smaller subset 
```{r}
sub <- genes [c(1:5, 55:60), c(1:5, 20:35, 55:60)]
```

con subset 1 (COPIAR G1)
```{r}
pheatmap(sub, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica")
```

con subset 2 -- DESPLEGAR VALORES
```{r}
pheatmap(sub, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica", fontsize = 8, annotation_legend = FALSE, display_numbers = TRUE, fontsize_number = 6)
```


Utilizar viridis, magma, plasma, cividis, inferno para diseño, asegurarse de instalar paqueteria y llamar a librería
install.packages("viridis")
library(viridis)
install.packages("viridisLite")
library(viridisLite)

Gráficos con color
```{r}
library(viridis)
pheatmap(sub, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica", fontsize = 8, annotation_legend = FALSE, display_numbers = TRUE, fontsize_number = 6, col = viridis_pal(option = "plasma") (6))

pheatmap(sub, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica", fontsize = 8, annotation_legend = FALSE, display_numbers = TRUE, fontsize_number = 6, col = viridis_pal(option = "magma") (6))

pheatmap(sub, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica", fontsize = 8, annotation_legend = FALSE, display_numbers = TRUE, fontsize_number = 6, col = viridis_pal(option = "viridis") (6))

pheatmap(sub, frontsize = 6, cluster_rows = T, cluster_cols = T, annotation_row = annotation_row, annotation_col = annotation_col, treeheight_row = 0, treeheight_col = 0, main = "expresión genetica", fontsize = 8, annotation_legend = FALSE, display_numbers = TRUE, fontsize_number = 6, col = viridis_pal(option = "inferno") (6))
```

Añadir elementos adicionales (identificar la distancia entre datos/genes)
```{r}
dist(sub)
```

Identificar el mapa de calor de la correlación de los datos (entre pacientes)
```{r}
pheatmap(cor(sub))
```

Identificar la correlación entre genes, utilizando matriz transpuesta
```{r}
trans <- t(sub)
pheatmap(cor(trans))
```


#####################################################################################################################

slightly artificial with color bar, without and with ordering:
cc <- rainbow(nrow(Ca))
heatmap(Ca, Rowv = FALSE, symm = TRUE, RowSideColors = cc, ColSideColors = cc,
        margins = c(6,6))
heatmap(Ca,		symm = TRUE, RowSideColors = cc, ColSideColors = cc,
        margins = c(6,6))
